

\indent I would like to thank the jury members that accepted the invitation 
to review this document and for their challenging questions which 
triggered interesting discussions during the defense. Their experience 
and varied background made this evaluation much richer. I want to thank 
particularly JC, who was the first to accept me for an internship in 
cosmology back in 2009.

I want to thank Dominique Fouchez and Cristinel Diaconu, 
for believing in my potential for research and teaching, and welcoming me 
in the cosmology group at the CPPM in 2020. 

I am only here today thanks to the trust of those that hired me in the past 
as a postdoc. 
Thank you, Kyle Dawson, for welcoming me in Utah - USA and showing me the guts of 
a spectroscopic survey and how important it is to understand all their pieces: 
from each bolt and nut to the cosmological constraints.  
Thank you, Will Percival, for hiring me at the ICG in Portsmouth - UK,
believing that I could lead research and supervise PhD students, and making me part of 
such a great group of cosmologists. I want to thank Florian Beutler and David Bacon 
for their experienced guidance during this time in England. 
 

Landing in such different countries and environments, and relearning how to adapt 
to new cultures was not an easy task. 
Thanks to a bunch of great friends, this task became easier. 
In Utah, I thank Antonio, Rafa, Amanda, Cari, Marc for all the adventures; 
the 5th Av crew: Kenneth, John, Alex, Jess, Hearin; 
and the folks from the football teams FC Breeze and Camaros FC. 
In Portsmouth, the weather was warmer thanks to many. First, Santi who had the best
idea ever to share a house. Our housemates: Jeanne, Sonia, Gui, Jose and our rockstar neighbour Liam. 
The French mafia: Coco, Daphné, Eroan, Nathalie, Bea, Andy, Églantine. 
And of course Davide, Eva, Hans, Rossana, Sesh \& Iza, from the ICG, making life look like 
an eternal (and fun) collaboration meeting.  
In Marseille, Tikago was the first to show me how fun this town can be. 
I feel lucky to have Ben as colleague in CPPM who made me  quickly feel like home.  
I also had the chance of having great colleagues becoming friends, such as 
Pauline, Dorothea, Chiara, Sylvain, Bianca. 
Thanks to Allyson, Laura, Natalia, Théo, Jess, Will, Chloé, Benoît, Ikram, Solo, 
for such great times we had (and those to come) in Marseille. 

Nothing I achieved would have been possible, nor I would have met these amazing people, 
if I did not have Caro by my side during most of that time. 
I am definitely thankful for following me all those years, while having to work double 
or triple than me in many aspects, like finding a job at every new location we would move. 
Unfortunately our paths had to diverge, but I will always be thankful for the love and affection 
from you and your family.  

Thanks Ondine for all the love and support this year. 
My life got beautiful new colors thanks to you and your energy. 

I dedicate this thesis to my whole family, particularly my parents Vilma and Hector, who are always there even if
my distance to you never got smaller since I left home. Being far away from you all is the toughest of the challenges
and time did not make it any easier. 
I dedicate this work specially to Tucho, Monica, and mi abuela. 
I will miss you forever.  
